---
title: "UCLA Nurse Blood Pressure Study"
subtitle: "CHL5222 Final Project (Group C)"
author: "Eileen Yu, Jianyue Bai, Ente Kang, Xiankai Yang, Sylvia Yin"
date: "2026-03-30"
output: 
  pdf_document:
    latex_engine: xelatex
---

**Study description:**
The UCLA Nurse Blood Pressure Study collected information from registered nurses in the Los Angeles area between 24 and 50 years of age on blood pressure (BP) and potential factors that contribute to hypertension. This information includes family history, and whether the subject had one or two hypertensive parents, as well as a wide range of measures of the physical and emotional condition of each nurse throughout the day. Researchers sought to study the links between BP and family history, personality, mood changes, working status, and menstrual phase.

The first BP measurement was taken half an hour before the subject’s normal start of work, and BP was then measured approximately every 20 minutes for the rest of the day. At each BP reading, the nurses also rate their mood on several dimensions, including how stressed they feel at the moment the BP is taken. In addition, the activity of each subject during the 10 minutes before each reading was measured using an actigraph worn on the waist.

**Research question:**
What are the factors that are associated with higher systolic blood pressure?

```{r setup, include=FALSE}
library(here)
library(PupillometryR)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(lme4)
library(lmerTest)
library(car)
library(performance)
library(lattice)
library(knitr)
library(broom.mixed)
# install.packages("mitools")
library(mitools)
library(mice)
library(naniar)
library(gtsummary)
library(kableExtra)
library(reshape2)
library(patchwork)
library(grid)
library(gridExtra)
knitr::opts_knit$set(root.dir = here())
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
theme_set(theme_bw())
```

# Variable Descriptions

| Variable | Description |
|---|---|
| `SNUM` | Unique subject identification number |
| `AGE` | Age in years |
| `DAY` | W: Workday or NW: Non-workday |
| `SYS` | Systolic blood pressure (mmHg) |
| `DIA` | Diastolic blood pressure (mmHg) |
| `HRT` | Heart rate (beats per minute) |
| `MNACT5` | Activity level (frequency of movements in 1-minute intervals over a 10-minute period) |
| `PHASE` | Menstrual phase (F: follicular; L: luteal) |
| `POSTURE` | Position during BP measurement (sitting, standing, or reclining) |
| `STR` | Self-rated stress level (1–5 scale, 5 = highest) |
| `HAP` | Self-rated happiness level (1–5 scale, 5 = highest) |
| `TIR` | Self-rated tiredness level (1–5 scale, 5 = highest) |
| `FH123` | Family history of hypertension (NO, YES, YESYES) |
| `time` | Time in minutes from midnight |
| `timept` | Measurement occasion (~50 per subject) |
| `timepass` | Time between measurements (minutes, starting at 0) |

# Read data
```{r}
nursebp <- read.csv(here("data","nursebp.csv"), stringsAsFactors = FALSE)
```

# Initial inspection
```{r}
str(nursebp)
```
```{r}
summary(nursebp)
```

```{r}
colSums(is.na(nursebp))
```

```{r}
dplyr::n_distinct(nursebp$SNUM) 
```

# Data cleaning
```{r}
nursebp <- nursebp %>%
  mutate(
    SNUM   = factor(SNUM),
    DAY    = factor(DAY),
    FH123  = factor(FH123),
    PHASE  = factor(PHASE),
    POSTURE = trimws(POSTURE)
  )

# clean POSTURE blank entries
nursebp$POSTURE[nursebp$POSTURE == ""] <- NA

# set posture levels explicitly
nursebp$POSTURE <- factor(
  nursebp$POSTURE,
  levels = c("RECLINE", "SIT", "STAND")
)

# check cleaned factors
str(nursebp)
colSums(is.na(nursebp))
```

# Exploratory data analysis
```{r subject-level descriptive table}
subj_data <- nursebp %>%
  group_by(SNUM) %>%
  summarise(
    AGE = first(AGE),
    FH123 = first(FH123),
    n_readings = n(),
    mean_SYS = mean(SYS, na.rm = TRUE),
    mean_HRT = mean(HRT, na.rm = TRUE),
    mean_STR = mean(STR, na.rm = TRUE),
    mean_HAP = mean(HAP, na.rm = TRUE),
    mean_TIR = mean(TIR, na.rm = TRUE),
    mean_MNACT5 = mean(MNACT5, na.rm = TRUE),
    
    .groups = "drop"
  ) %>%
  mutate(
    FH123 = factor(FH123)
  )

tbl_subj <- subj_data %>%
  select(
    AGE, FH123,
    n_readings,
    mean_SYS,
    mean_HRT,
    mean_STR, mean_HAP, mean_TIR,
    mean_MNACT5
  ) %>%
  tbl_summary(
    missing = "no",
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    label = list(
      AGE ~ "Age (years)",
      FH123 ~ "Family history",
      n_readings ~ "Number of BP readings",
      mean_SYS ~ "Mean systolic BP (mmHg)",
      mean_HRT ~ "Mean heart rate",
      mean_STR ~ "Mean stress level",
      mean_HAP ~ "Mean happiness",
      mean_TIR ~ "Mean tiredness",
      mean_MNACT5 ~ "Mean activity level"
    )
  ) %>%
  bold_labels()

tbl_subj
```

```{r sys-histogram}
library(ggplot2)

ggplot(nursebp, aes(x = SYS)) +
  geom_histogram(
    bins = 20,
    color = "black",
    fill = "skyblue"
  ) +
  labs(
    title = "Distribution of Systolic Blood Pressure",
    x = "Systolic Blood Pressure (mmHg)",
    y = "Count"
  ) +
  theme_minimal(base_size = 14)
```

```{r sys-time-subject}
nursebp$hour <- nursebp$time / 60
ggplot(nursebp, aes(x = hour, y = SYS, group = SNUM)) +
  geom_line(alpha = 0.1) +
  geom_smooth(aes(group = 1), method = "loess", linewidth = 1) +
  theme_minimal() +
  labs(
    title = "Within-Subject Blood Pressure Patterns Over Time",
    x = "Time (hours)",
    y = "Systolic Blood Pressure (mmHg)"
  )
```

```{r population trend vs time from midnight}
ggplot(nursebp, aes(x = time, y = SYS)) +
    geom_smooth(method = "loess", se = TRUE) +
    labs(
        title = "Smoothed Trend of Systolic Blood Pressure",
        x = "Time (minutes from midnight)",
        y = "Systolic Blood Pressure (mmHg)"
    ) +
    theme_minimal(base_size = 12)
```

```{r Correlation and association among variables}
# Continuous-variable correlation check 
cont_vars <- nursebp[, c("SYS", "DIA", "HRT", "MNACT5", "time", "STR", "HAP", "TIR")]
cor(cont_vars, use = "complete.obs")

cor(nursebp[, c("SYS", "DIA")], 
    use = "complete.obs")
cor(nursebp[, c("STR", "HAP", "TIR")], 
    use = "complete.obs")
```


```{r kept vs removed comparison }
# complete cases

vars <- c("SNUM","SYS","time","HRT","MNACT5","PHASE","AGE","FH123", "STR", "HAP", "TIR", "DAY", "POSTURE")

dat_gls <- nursebp[complete.cases(nursebp[, vars]), ]

#After excluding observations with missing values in key predictors, 7877 out of 9573 observations (82.3%) were retained.

nursebp$kept <- ifelse(
    !is.na(nursebp$STR) &
        !is.na(nursebp$TIR) &
        !is.na(nursebp$HAP) &
        !is.na(nursebp$MNACT5)&
        !is.na(nursebp$POSTURE),
    "Kept", "Removed"
)

nursebp %>%
    group_by(kept) %>%
    summarise(
        SYS_mean = mean(SYS, na.rm = TRUE),
        HRT_mean = mean(HRT, na.rm = TRUE),
        AGE_mean = mean(AGE, na.rm = TRUE)
    )

tab <- table(nursebp$FH123, nursebp$kept)

# Column % (inside Kept / Removed) FH123
pct_col <- round(prop.table(tab, margin = 2) * 100, 1)

# Row % (Removed rate)
pct_removed <- round(tab[, "Removed"] / rowSums(tab) * 100, 1)

# Combine everything
result_fh <- data.frame(
  FH123 = rownames(tab),
  Kept = paste0(tab[, "Kept"], " (", pct_col[, "Kept"], "%)"),
  Removed = paste0(tab[, "Removed"], " (", pct_col[, "Removed"], "%)"),
  `% Removed` = paste0(pct_removed, "%")
)

result_fh

# Day

tab_day <- table(nursebp$DAY, nursebp$kept)

# Column % (inside Kept / Removed)
pct_col <- round(prop.table(tab_day, margin = 2) * 100, 1)

# Row % (% Removed within each DAY group)
pct_removed <- round(tab_day[, "Removed"] / rowSums(tab_day) * 100, 1)

# Combine
result_day <- data.frame(
  DAY = rownames(tab_day),
  Kept = paste0(tab_day[, "Kept"], " (", pct_col[, "Kept"], "%)"),
  Removed = paste0(tab_day[, "Removed"], " (", pct_col[, "Removed"], "%)"),
  `% Removed` = paste0(pct_removed, "%")
)

result_day

p_result_day<-wrap_elements(tableGrob(result_day))

p_result_fh<-wrap_elements(tableGrob(result_fh))

p_sys<-ggplot(nursebp, aes(x = kept, y = SYS)) +
    geom_boxplot(fill = "lightblue") +
    #labs(title = "SYS: Kept vs Removed") +
    labs(y="Systolic Blood Pressure(SYS)")+
    theme_minimal()

p_hrt <- ggplot(nursebp, aes(x = kept, y = HRT, fill = kept)) +
  geom_boxplot() +
  labs( x = NULL, y = "Heart Rate") +#title = "HRT: Kept vs Removed",
  theme_minimal() +
  theme(legend.position = "none")

p_age <- ggplot(nursebp, aes(x = kept, y = AGE, fill = kept)) +
  geom_boxplot() +
  labs(x = NULL, y = "Age") +#title = "AGE: Kept vs Removed"
  theme_minimal() +
  theme(legend.position = "none")
p_sys
p_hrt
p_age

(p_result_day+ggtitle("Workday or Non-Workday")) /(p_result_fh+ggtitle("FH123"))

```


```{r Preliminary fixed-effect collinearity assessment}
# Exclude DIA from main model for substantive interpretability:
# DIA is another BP measure and may dominate explanation of SYS.
vif_model <- lm(
  SYS ~ AGE + DAY + FH123 + PHASE + POSTURE +
    HRT + MNACT5 + STR + HAP + TIR + hour,
  data = nursebp
)

car::vif(vif_model)
```

# Methods

A. Linear Mixed-Effects Model

We modeled systolic blood pressure (SYS) using linear mixed-effects models, accounting for nurses' repeated measurements. Time was measured in hours since midnight. Since the initial plots showed a gentle curve over time rather than a straight line, we tested both linear and quadratic time terms to see which fit the data better.

Time construction
```{r}
nursebp <- nursebp %>%
  mutate(
    hour = time / 60,
    hour_c = hour - mean(hour, na.rm = TRUE),
    hour2_c = hour_c^2
  )
```

Model selection

We first compared random intercept and random slope models, then focused on linear and quadratic time specifications under the selected random-effects structure.
```{r}
# random intercept model
m_intercept <- lmer(
  SYS ~ hour_c + hour2_c +
    AGE + DAY + FH123 + PHASE + POSTURE +
    STR + HAP + TIR + HRT + MNACT5 +
    (1 | SNUM),
  data = nursebp,
  REML = FALSE
)

# random slope model
m_random_slope <- lmer(
  SYS ~ hour_c + hour2_c +
    AGE + DAY + FH123 + PHASE + POSTURE +
    STR + HAP + TIR + HRT + MNACT5 +
    (1 + hour_c | SNUM),
  data = nursebp,
  REML = FALSE,
  control = lmerControl(
    optimizer = "bobyqa",
    optCtrl = list(maxfun = 1e5)
  )
)

AIC(m_intercept, m_random_slope)
BIC(m_intercept, m_random_slope)
```
```{r}
anova(m_intercept, m_random_slope)
VarCorr(m_random_slope)
```
```{r}
m_linear_time <- lmer(
  SYS ~ hour_c +
    AGE + DAY + FH123 + PHASE + POSTURE +
    STR + HAP + TIR + HRT + MNACT5 +
    (1 + hour_c | SNUM),
  data = nursebp,
  REML = FALSE,
  control = lmerControl(
    optimizer = "bobyqa",
    optCtrl = list(maxfun = 1e5)
  )
)

m_quadratic_time <- lmer(
  SYS ~ hour_c + hour2_c +
    AGE + DAY + FH123 + PHASE + POSTURE +
    STR + HAP + TIR + HRT + MNACT5 +
    (1 + hour_c | SNUM),
  data = nursebp,
  REML = FALSE,
  control = lmerControl(
    optimizer = "bobyqa",
    optCtrl = list(maxfun = 1e5)
  )
)

AIC(m_linear_time, m_quadratic_time)
BIC(m_linear_time, m_quadratic_time)
```
```{r}
anova(m_linear_time, m_quadratic_time)
```
Final model

Ultimately, we selected the model with a random slope for time and both linear and quadratic time terms.
```{r}
final_model <- m_quadratic_time
summary(final_model)

car::linearHypothesis(final_model, "hour2_c = 0")
r2_vals <- r2_nakagawa(final_model)
r2_vals
```
```{r}
fe_summary <- summary(final_model)$coefficients
plot_df <- data.frame(
  term = rownames(fe_summary),
  estimate = fe_summary[, "Estimate"],
  std_error = fe_summary[, "Std. Error"],
  p_value = fe_summary[, "Pr(>|t|)"]
)

plot_df <- plot_df %>%
  mutate(
    conf_low = estimate - 1.96 * std_error,
    conf_high = estimate + 1.96 * std_error
  ) %>%
  filter(term != "(Intercept)")

plot_df
```

```{r}
plot_df$term <- factor(plot_df$term, levels = rev(plot_df$term))

ggplot(plot_df, aes(x = estimate, y = term)) +
  geom_point() +
  geom_errorbarh(aes(xmin = conf_low, xmax = conf_high), height = 0.2) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    title = "Fixed Effects Estimates from the Final Model",
    x = "Estimate (95% CI)",
    y = NULL
  ) +
  theme_minimal()
```

Model diagnostics
```{r}
# Residual vs fitted
plot(fitted(final_model), resid(final_model),
     main = "Residuals vs Fitted Values",
     xlab = "Fitted values",
     ylab = "Residuals",
     pch = 16)
abline(h = 0, col = "red", lty = 2)
```

```{r}
qqnorm(residuals(final_model))
qqline(residuals(final_model))
```

Model fit statistics
```{r}
model_fit <- data.frame(
  AIC = round(AIC(final_model), 2),
  BIC = round(BIC(final_model), 2),
  Marginal_R2 = round(r2_vals$R2_marginal, 3),
  Conditional_R2 = round(r2_vals$R2_conditional, 3)
)

model_fit
```
```{r}
fe_summary <- summary(final_model)$coefficients
plot_df <- data.frame(
  term = rownames(fe_summary),
  estimate = fe_summary[, "Estimate"],
  std_error = fe_summary[, "Std. Error"],
  p_value = fe_summary[, "Pr(>|t|)"]
)

plot_df <- plot_df %>%
  mutate(
    conf_low = estimate - 1.96 * std_error,
    conf_high = estimate + 1.96 * std_error
  ) %>%
  filter(term != "(Intercept)")

plot_df
```

```{r}
plot_df$term <- factor(plot_df$term, levels = rev(plot_df$term))

ggplot(plot_df, aes(x = estimate, y = term)) +
  geom_point() +
  geom_errorbarh(aes(xmin = conf_low, xmax = conf_high), height = 0.2) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    title = "Fixed Effects Estimates from the Final Model",
    x = "Estimate (95% CI)",
    y = NULL
  ) +
  theme_minimal()
```

# Sensitivity analysis: multiple imputation

Because several predictors contained missing values, multiple imputation was used as a sensitivity analysis under a missing at random assumption.
```{r}
na_cols <- c("MNACT5", "POSTURE", "STR", "HAP", "TIR")
colSums(is.na(nursebp[, na_cols]))

vars_for_imp <- c(
  "SYS", "AGE", "HRT", "MNACT5", "STR", "HAP", "TIR",
  "DAY", "PHASE", "POSTURE", "FH123",
  "hour_c", "hour2_c", "SNUM"
)

df_sub <- nursebp[, vars_for_imp]

method <- mice::make.method(df_sub)
pred <- mice::make.predictorMatrix(df_sub)

method["SNUM"] <- ""
pred[, "SNUM"] <- 0
pred["SNUM", ] <- 0

method[c("SYS","AGE","HRT","MNACT5","STR","HAP","TIR", "hour_c", "hour2_c")] <- "pmm"
method[c("DAY","PHASE")] <- "logreg"
method[c("POSTURE","FH123")] <- "polyreg"
```


```{r}
imp_data <- mice(
  df_sub,
  m = 5,
  method = method,
  maxit = 20,
  predictorMatrix = pred,
  seed = 406,
  printFlag = FALSE
)

fit_lme_mi <- with(
  imp_data,
  lmer(
    SYS ~ AGE + HRT + STR + HAP + TIR +
      DAY + POSTURE + PHASE + FH123 +
      hour_c + hour2_c + MNACT5 +
      (1 + hour_c | SNUM)
  )
)

pooled_lme <- pool(fit_lme_mi)
summary(pooled_lme)
```





