---
title: "Correlated_Group_Project"
format: html
editor: visual
---

```{r}
df <- read.csv('nursebp.csv')
colnames(df)
```

```{r}
df$SNUM <- factor(df$SNUM)
df$SYS <- as.numeric(df$SYS)
df$DIA <- as.numeric(df$DIA)
df$HRT <- as.numeric(df$HRT)
df$MNACT5 <- as.numeric(df$MNACT5)
df$PHASE <- factor(df$PHASE)
df$DAY <- factor(df$DAY)
df$POSTURE <- factor(df$POSTURE)
df$STR <- factor(df$STR)
df$HAP <- factor(df$HAP)
df$TIR <-factor(df$TIR)
df$AGE <- as.numeric(df$AGE)
df$FH123 <-factor(df$FH123)
df$time <- as.numeric(df$time)
df$timepass <- as.numeric(df$timepass)
df$timept <- factor(df$timept)

# Convert all empty strings and single spaces to true NA
df[df == "" | df == " "] <- NA
```

## **Missing Pattern**

```{r}
library(naniar)

# Visualizes where the NAs are across your variables
vis_miss(df)

# Shows the intersection of missingness (e.g., who is missing both SYS and HRT?)
gg_miss_upset(df)
```

Eileen: logistic regression to assess . not MCAR. it is MAR.

## **Systolic Blood Pressure (SYS) Over Time**

```{r}
library(ggplot2)

ggplot(df, aes(x = time, y = SYS, group = SNUM, color = as.factor(SNUM))) +
  geom_line(alpha = 0.4) +
  # Adding a trend line to see the 'average' minute-by-minute change
  geom_smooth(aes(group = 1), method = "gam", color = "black", size = 1.2) +
  labs(title = "SYS Change Over Minutes",
       x = "Minutes (from midnight/baseline)",
       y = "Systolic BP") +
  theme_minimal() +
  theme(legend.position = "none")
```

## Distribution of SYS by Categorical Predictors (Violin Plots)

```{r}
library(ggplot2)

# 1. Identify categorical columns (excluding SNUM/ID and the target SYS)
# Adjust the 'setdiff' list to exclude any columns you don't want to plot
cat_vars <- names(df)[sapply(df, function(x) is.factor(x) || is.character(x))]
cat_vars <- setdiff(cat_vars, c("SNUM", "ID")) 

# 2. Loop through and create plots
for (var in cat_vars) {
  
  p <- ggplot(df, aes(x = .data[[var]], y = SYS, fill = .data[[var]])) +
    geom_violin(trim = FALSE, alpha = 0.7) +
    geom_boxplot(width = 0.1, color = "black", outlier.shape = NA) + # Adds a mini-boxplot inside
    labs(title = paste("Distribution of SYS by", var),
         x = var,
         y = "Systolic BP") + 
    theme_minimal() +
    theme(legend.position = "none")
  
  print(p)
}
```

## Distribution of SYS by Continuous Predictors (Violin Plots)

```{r}
library(ggplot2)
library(dplyr)

# 1. Identify Numeric Predictors (excluding your outcome and IDs)
num_vars <- names(df)[sapply(df, is.numeric)]
num_vars <- setdiff(num_vars, c("SYS", "SNUM"))

# 2. Create a temporary plotting dataframe
df_plot <- df

# 3. Bin numeric variables and plot
for (num_var in num_vars) {
  
  # Create 4 bins based on quantiles
  # We use factor() to ensure the plot treats them as categories
  df_plot[[paste0(num_var, "_binned")]] <- factor(
    ntile(df_plot[[num_var]], 8), 
    labels = c("Q1 (Lowest)", "Q2", "Q3", "Q4", "Q5", "Q6", "Q7", "Q8")
  )
  
  # Define the new binned variable name
  current_var <- paste0(num_var, "_binned")
  
  # Remove NAs for the specific variable being plotted
  plot_data <- df_plot[!is.na(df_plot[[current_var]]), ]
  
  # Generate the plot
  p <- ggplot(plot_data, aes(x = .data[[current_var]], y = SYS, fill = .data[[current_var]])) +
    geom_violin(trim = FALSE, alpha = 0.6) +
    geom_boxplot(width = 0.1, color = "black", outlier.shape = NA) +
    scale_fill_brewer(palette = "Blues") + # Clean blue color scale
    labs(title = paste("SYS Distribution by", num_var, "(Binned)"),
         subtitle = "Variable split into equal-sized quartiles",
         x = paste(num_var, "Quartiles"),
         y = "Systolic BP") +
    theme_minimal() +
    theme(legend.position = "none")
  
  print(p)
}
```

Eileen: DIA is too strongly correlated with SYS. So perhaps delete it from predictor variables.

## **Collinearity Matrix of Predictors**

```{r}
library(corrplot)
library(dplyr)

# 1. Prepare the data
# Ensure categories are factors and self-ratings are numeric
df_test <- df %>%
  mutate(
    # Convert mood/stress to numeric if they aren't already
    across(c(STR, HAP, TIR), ~as.numeric(as.character(.))),
    # Ensure categorical variables are factors
    across(c(PHASE, POSTURE, DAY), as.factor)
  ) 

# 2. Expand Factors into Dummy Variables (0/1)
# model.matrix creates a column for each level (e.g., POSTUREStanding, POSTUREReclining)
# we remove the intercept [,-1]
options(na.action='na.pass') # Allow model.matrix to keep rows with NAs

meta_matrix <- model.matrix(~ 1 + STR + HAP + TIR + MNACT5 + AGE + time + 
                              PHASE + POSTURE + DAY + FH123 + DIA + HRT, 
                            data = df_test)

meta_matrix <- meta_matrix[, colnames(meta_matrix) != "(Intercept)"]

# 3. Generate Correlation Matrix
# This now includes specific levels like PHASEL and POSTUREStanding
cor_matrix_full <- cor(meta_matrix, use = "pairwise.complete.obs")

# 4. Plot with adjusted labels for a larger matrix
corrplot(cor_matrix_full, 
         method = "color", 
         addCoef.col = "black", 
         number.cex = 0.6,    # Smaller numbers to fit
         tl.cex = 0.7,        # Smaller text labels
         tl.col = "black",
         title = "Full Predictor Collinearity Check (Numeric & Dummy)",
         mar = c(0,0,1,0))

```

## **Interaction Effects Check**

```{r}
library(ggplot2)
library(patchwork)
library(dplyr)

# 1. Setup Theme and Palette
clean_theme <- theme_minimal(base_size = 9) + 
               theme(panel.grid.minor = element_blank(),
                     legend.title = element_text(face = "bold", size = 7),
                     legend.text = element_text(size = 6),
                     plot.title = element_text(size = 9, face = "bold"),
                     axis.title = element_text(size = 8))

# 5-level warm palette for Stress (Yellow to Dark Red)
warm_5 <- c("1" = "#ffffb2", "2" = "#fecc5c", "3" = "#fd8d3c", "4" = "#f03b20", "5" = "#bd0026")

l_width <- 0.6 
p_size  <- 1.2

# --- 1. POSTURE * AGE (Binned Age) ---
p1 <- ggplot(df, aes(x = POSTURE, y = SYS, color = cut_number(AGE, 3), group = cut_number(AGE, 3))) +
  stat_summary(fun = mean, geom = "point", size = p_size) +
  stat_summary(fun = mean, geom = "line", linewidth = l_width) +
  scale_color_brewer(palette = "YlOrRd", name = "Age Group") +
  labs(title = "Postural BP x Age", x = "Posture", y = "SYS") +
  clean_theme

# --- 2. STR * TIME (Time-Varying Recovery) ---
# Switched to GAM to see non-linear recovery curves across all 5 STR levels
p2 <- ggplot(df, aes(x = time, y = SYS, color = STR)) +
  geom_smooth(method = "gam", formula = y ~ s(x, bs = "cs"), se = FALSE, linewidth = l_width) +
  scale_color_manual(values = warm_5, name = "Stress (1-5)") +
  labs(title = "Time x Stress", x = "Minutes", y = "SYS") +
  clean_theme

# --- 3. STR * HAP (5-level Stress) ---
p3 <- ggplot(df, aes(x = STR, y = SYS, color = HAP, group = HAP)) +
  stat_summary(fun = mean, geom = "point", size = p_size) +
  stat_summary(fun = mean, geom = "line", linewidth = l_width) +
  scale_color_brewer(palette = "Greens", name = "Happiness") +
  labs(title = "Buffering: Stress x Happiness", x = "Stress Level", y = "SYS") +
  clean_theme

# --- 4. TIR * MNACT5 (Activity Interaction) ---
p4 <- ggplot(df, aes(x = TIR, y = SYS, color = cut_number(MNACT5, 2), group = cut_number(MNACT5, 2))) +
  stat_summary(fun = mean, geom = "point", size = p_size) +
  stat_summary(fun = mean, geom = "line", linewidth = l_width) +
  scale_color_manual(values = c("#74a9cf", "#ce1256"), name = "Activity") +
  labs(title = "Tiredness x Activity", x = "Tiredness", y = "SYS") +
  clean_theme

# --- 5. PHASE * STR (Hormonal Stress Reactivity) ---
p5 <- ggplot(df, aes(x = STR, y = SYS, color = PHASE, group = PHASE)) +
  stat_summary(fun = mean, geom = "point", size = p_size) +
  stat_summary(fun = mean, geom = "line", linewidth = l_width) +
  # Using distinct colors for Follicular and Luteal phases
  scale_color_manual(values = c("F" = "#66c2a5", "L" = "#fc8d62"), 
                     name = "Menstrual Phase",
                     labels = c("F" = "Follicular", "L" = "Luteal")) +
  labs(title = "Stress Reactivity by Menstrual Phase", 
       x = "Stress Level", 
       y = "SYS") +
  clean_theme +
  guides(color = guide_legend(override.aes = list(linewidth = 1)))

# Combine
(p1 | p2) / (p3 | p4) / p5
```

## Standardization of Continuous Variables

```{r}
library(dplyr)

df_test <- df_test %>%
  mutate(
    # This z-scores all numeric columns EXCEPT snum and time
    across(
      .cols = where(is.numeric) & !c(SNUM), 
      .fns = ~ as.numeric(scale(.)),
      .names = "{.col}_z"  # This creates new names like AGE_z, SYS_z, etc.
    )
  )

df_test$SNUM <- factor(df_test$SNUM)
df_test$SYS <- as.numeric(df_test$SYS)
df_test$DIA <- as.numeric(df_test$DIA)
df_test$HRT <- as.numeric(df_test$HRT)
df_test$MNACT5 <- as.numeric(df_test$MNACT5)
df_test$PHASE <- factor(df_test$PHASE)
df_test$DAY <- factor(df_test$DAY)
df_test$POSTURE <- factor(df_test$POSTURE)
df_test$STR <- factor(df_test$STR)
df_test$HAP <- factor(df_test$HAP)
df_test$TIR <-factor(df_test$TIR)
df_test$AGE <- as.numeric(df_test$AGE)
df_test$FH123 <-factor(df_test$FH123)
df_test$time <- as.numeric(df_test$time)
df_test$timepass <- as.numeric(df_test$timepass)
df_test$timept <- factor(df_test$timept)

df_test$SNUM <- factor(df_test$SNUM)
df_test$SYS_z <- as.numeric(df_test$SYS_z)
df_test$DIA_z <- as.numeric(df_test$DIA_z)
df_test$HRT_z <- as.numeric(df_test$HRT_z)
df_test$MNACT5_z <- as.numeric(df_test$MNACT5_z)
df_test$PHASE <- factor(df_test$PHASE)
df_test$DAY <- factor(df_test$DAY)
df_test$POSTURE <- factor(df_test$POSTURE)
df_test$STR <- factor(df_test$STR)
df_test$HAP <- factor(df_test$HAP)
df_test$TIR <-factor(df_test$TIR)
df_test$AGE_z <- as.numeric(df_test$AGE_z)
df_test$FH123 <-factor(df_test$FH123)
df_test$time_z <- as.numeric(df_test$time_z)
df_test$timepass_z <- as.numeric(df_test$timepass_z)
df_test$timept <- factor(df_test$timept)
```

```{r}
ggplot(df_test, aes(x = time_z, y = SYS, group = SNUM, color = as.factor(SNUM))) +
  geom_line(alpha = 0.4) +
  # Adding a trend line to see the 'average' minute-by-minute change
  geom_smooth(aes(group = 1), method = "gam", color = "black", size = 1.2) +
  labs(title = "SYS Change Over Minutes",
       x = "Minutes (from midnight/baseline)",
       y = "Systolic BP") +
  theme_minimal() +
  theme(legend.position = "none")
```

## Spline GEE

```{r}
library(geepack)

vars <- c("SYS", "time", "STR", "PHASE", "POSTURE", "HRT",
          "AGE", "HAP", "TIR", "MNACT5", "DAY", "FH123", "SNUM")

df_clean <- df_test[
  complete.cases(df_test[, vars]),
]

# Convert problematic entries to NA first
df_test$POSTURE[df_test$POSTURE %in% c("NA", "", " ")] <- NA

# Then remove incomplete cases
df_clean <- df_test[
  complete.cases(df_test[, vars]),
]

# Drop unused levels
df_clean$POSTURE <- droplevels(df_clean$POSTURE)

# Create time knot
df_clean$time_knot <- pmax(0, df_clean$time_z - 1)

gee_piecewise_ar <- geeglm(
  SYS ~ time_z + time_knot + HRT_z + FH123 + STR + PHASE +
        POSTURE + AGE_z + HAP + TIR + MNACT5_z + DAY,
  data = df_clean,
  id = SNUM,
  family = gaussian,
  corstr = "ar1"
)
summary(gee_piecewise_ar)
```

```{r}
# Extract fitted values and Pearson residuals
fitted_vals <- predict(gee_piecewise_ar)
resids <- residuals(gee_piecewise_ar, type = "pearson")

# Create the plot
my_colour <- adjustcolor("steelblue", alpha.f = 0.2)
plot(fitted_vals, resids, 
     xlab = "Fitted Values", 
     ylab = "Pearson Residuals",
     main = "Residual vs. Fitted Plot (GEE)",
     pch = 20, col = my_colour)
abline(h = 0, lty = 2, col = "red") # Add a horizontal line at 0
```

## Model Comparison

```{r}
# Standard Linear Model (No Knot)
gee_linear_ar <- geeglm(
  SYS ~ time_z + HRT_z + FH123 + STR + PHASE + 
        POSTURE + AGE_z + HAP + TIR + MNACT5_z + DAY,
  data = df_clean, 
  id = SNUM, 
  family = gaussian, 
  corstr = "ar1"
)
```

```{r}
# Piecewise Model with Exchangeable Correlation
gee_piecewise_exch <- geeglm(
  SYS ~ time_z + time_knot + HRT_z + FH123 + STR + PHASE + 
        POSTURE + AGE_z + HAP + TIR + MNACT5_z + DAY,
  data = df_clean, 
  id = SNUM, 
  family = gaussian, 
  corstr = "exchangeable"
)
```

```{r}
# Standard Linear Model (No Knot)
gee_linear_exch <- geeglm(
  SYS ~ time_z + HRT_z + FH123 + STR + PHASE + 
        POSTURE + AGE_z + HAP + TIR + MNACT5_z + DAY,
  data = df_clean, 
  id = SNUM, 
  family = gaussian, 
  corstr = "exchangeable"
)
```

```{r}
# 1. Fit the final winning model (Linear Exchangeable)
gee_final <- geeglm(
  SYS ~ time_z + HRT_z + FH123 + STR + PHASE + 
        POSTURE + AGE_z + HAP + TIR + MNACT5_z + DAY + PHASE:POSTURE,
  data = df_clean, 
  id = SNUM, 
  family = gaussian, 
  corstr = "exchangeable"
)

summary(gee_final)
```


```{r}
library(ggplot2)
library(gridExtra)

# 1. Extract the required components from your GEE model
df_clean$pearson_res <- residuals(gee_linear_exch, type = "pearson")
all_coefs <- coef(gee_linear_exch)

# 2. Function to generate a Partial Residual Plot
plot_linearity <- function(var_name, title_label) {
  # Calculate partial residuals: (Beta * X) + Residuals
  beta_val <- all_coefs[var_name]
  df_clean$partial_res <- (beta_val * df_clean[[var_name]]) + df_clean$pearson_res
  
  ggplot(df_clean, aes_string(x = var_name, y = "partial_res")) +
    geom_point(alpha = 0.2, color = "grey70") +
    # The 'Truth': Smoothed Loess line
    geom_smooth(method = "loess", color = "blue", se = FALSE, size = 1) + 
    # The 'Assumption': Your model's linear slope
    geom_abline(intercept = 0, slope = beta_val, color = "red", linetype = "dashed", size = 1) +
    labs(title = paste("Linearity Check:", title_label),
         y = "Partial Residuals", x = var_name) +
    theme_minimal()
}

# 3. Generate the plots
p1 <- plot_linearity("time_z", "Time (std)")
p2 <- plot_linearity("HRT_z", "Heart Rate (std)")
p3 <- plot_linearity("AGE_z", "Age (std)")
p4 <- plot_linearity("MNACT5_z", "Activity (std)")

# 4. Display in a grid
grid.arrange(p1, p2, p3, p4, ncol = 2)
```

```{r}
compute_QIC <- function(model) {
  # 1. Ensure we have the data and basic components
  y <- model$y
  mu <- model$fitted.values
  n <- length(y)
  
  # 2. Calculate Quasi-Likelihood for Gaussian (Normal) family
  # For Gaussian, Q is -0.5 * RSS (Residual Sum of Squares)
  rss <- sum((y - mu)^2)
  quasi_lik <- -0.5 * rss
  
  # 3. Fit an independence model to get the Inverse model-based covariance (AI)
  # We use update() to ensure all other parameters stay the same
  model_indep <- update(model, corstr = "independence")
  
  # 4. Extract Covariance Matrices
  # Vr = Robust (sandwich) variance-covariance matrix
  # Vi = Model-based variance-covariance matrix (under independence)
  Vr <- model$geese$vbeta
  Vi <- model_indep$geese$vbeta
  
  # 5. Calculate the Trace component
  # solve(Vi) gives us the AI matrix
  AI <- solve(Vi)
  trace_val <- sum(diag(AI %*% Vr))
  
  # 6. Compute QIC and QICu
  # QIC = -2Q + 2*trace
  # QICu = -2Q + 2*p (where p is number of parameters)
  p <- length(model$coefficients)
  qic <- -2 * quasi_lik + 2 * trace_val
  qicu <- -2 * quasi_lik + 2 * p
  
  return(data.frame(QIC = qic, QICu = qicu, Trace = trace_val, p = p))
}
```

```{r}
# Function to calculate Root Mean Square Error for GEE
compute_rmse <- function(model) {
  # For Gaussian GEE, we can use the difference between observed and predicted
  preds <- predict(model)
  obs <- model$y
  rmse_val <- sqrt(mean((obs - preds)^2))
  return(round(rmse_val, 3))
}
```

```{r}
# 1. Run the RMSE function for each model
rmse1 <- compute_rmse(gee_piecewise_ar)
rmse2 <- compute_rmse(gee_linear_ar)
rmse3 <- compute_rmse(gee_piecewise_exch)
rmse4 <- compute_rmse(gee_linear_exch)

# 2. Collect the QIC values (assuming you used a QIC function)
qic1 <- compute_QIC(gee_piecewise_ar)
qic2 <- compute_QIC(gee_linear_ar)
qic3 <- compute_QIC(gee_piecewise_exch)
qic4 <- compute_QIC(gee_linear_exch)
```


```{r}
library(knitr)
library(kableExtra)

model_results <- data.frame(
  Model = c("Piecewise AR1", 
            "Linear AR1", 
            "Piecewise Exchange", 
            "Linear Exchange"),
  QIC = c(qic1$QIC, qic2$QIC, qic3$QIC, qic4$QIC),
  RMSE = c(rmse1, rmse2, rmse3, rmse4), # Example values; plug in your results here
  Parameters = c(
    ncol(model.matrix(gee_piecewise_ar)) + length(gee_piecewise_ar$geese$alpha),
    ncol(model.matrix(gee_linear_ar)) + length(gee_linear_ar$geese$alpha),
    ncol(model.matrix(gee_piecewise_exch)) + length(gee_piecewise_exch$geese$alpha),
    ncol(model.matrix(gee_linear_exch)) + length(gee_linear_exch$geese$alpha)
  ))

# 2. Format the table for your QMD/Presentation
model_results %>%
  kbl(caption = "Table 1: GEE Model Fit and Predictive Accuracy Metrics",
      format = "html", 
      align = "lccc",
      col.names = c("Model Structure", "QIC", "RMSE (mmHg)", "No. Params")) %>%
  kable_classic(full_width = F, html_font = "Arial") %>%
  row_spec(1, bold = T, background = "lightblue") 
```

